---
title: 'Comparative methods for biological surface patterns in R: pholidosis'
author: "Isaac Krone"
date: "2025-01-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# pholidosis

This vignette demonstrates the functionality of the ```pholidosis``` package, which implements a novel methodology for comparing biological surface patterns with clear homology relationships. The primary goal of ```pholidosis``` is to construct  an edit distance matrix that measures the difference between patterns.  

The document below provides a straightforward demonstration of the ```pholidosis``` workflow, beginning with an excel workbook and ending with a Principal Coordinates analysis based on the edit distance matrix for a set of head scalation patterns of dibamid lizards (family Dibamidae).

Future releases will provide more detailed insight into the package's functions and should provide some help in troubleshooting when you want to use your own data. 

## 1 Installation
```pholidosis``` makes heavy use of the ```igraph``` R package. See <https://r.igraph.org/> for instructions on installing this package, which may require installing dependencies outside of R.

```pholidosis``` can be installed from GitHub with:
```{r eval = F}
#install.packages("remotes")
remotes::install_github("dibamus/pholidosis")
```
Functions in ```pholidosis``` use various dependencies:

```{r eval =F}
library('dplyr')
library('igraph')
library ('ggraph')
library('readxl')
library('stringr')
library('tibble')
library('tidygraph')
library('tidyverse')
library('utils')
```

## 2 Quick start example

Let's walk through using ```pholidosis``` to import a set of adjacency matrices, visualize their corresponding graphs, and produce a distance matrix.

To run this tutorial, you'll need the following packages loaded:
```{r packages}
library('igraph')
library('pholidosis')
library('readxl') 
library('ggraph') # for plotting
library('tidygraph') # for plotting
```
### 2.1 Import and plot
With those loaded up, we'll need to find the Dibamids demo dataset included in the ```pholidosis``` package. It's an .xlsx file, so we need to find the file path for it before we load it. Then we'll use the ```pholidosis::excel_to_network()``` function to load in these adjacency matrices as a list of networks.
```{r loadin}
pth <- system.file("extdata", "DibamidaeDemo.xlsx", package = "pholidosis")
ntw <- excel_to_network(pth)
```

Let's take a quick look at a graph to see what it looks like.

```{r plot, warning = FALSE, message = FALSE}

pholidosis_plot(ntw[[1]], title = "Anelytropsis papillosus \npholidosis network")
```

In this visualization, we can see the topology of the pholidosis network laid flat, as if you were viewing a lizard's head from above. The front of the lizard's head is on the right side, and the rest of the lead is laid out towards the left. Vertices corresponding to the scales of the lower jaw are laid out further to the left. Here, vertices are color coded by the scale type they represent. Numbers appear in vertices where those vertices represent the same type of scale (for instance, a supralabial). The bold lines between the rostral and internarial vertices indicate a higher edge weight; those scales are fused together in this animal.
Note that the automatic layout method isn't perfect; graphs will often be "stretched" or "folded" rather than laid out totally flat.

### 2.2 Measure graph edit distance

Assuming all of your graphs are topologically valid and don't include any difficult graph transformations that the algorithm can't handle, creating a distance distance matrix is easy, though not necessarily quick. This may take a few minutes to run.

```{r distance matrix}
dist <- netDistMat(ntw)
```

Note that the matrix is three dimensional, with a size of n x n x 3, where n is the length of the list of graphs input. There is an n x n matrix for each of 3 components: *w*, the number of vertex differences between the networks; *e*, the number of edge differences between the networks, and *f*, the number of edge weight differences between the networks.

Separating out these components can help you understand how these different features are contributing to the diversity of the system you're studying, but the measurements contained in them are equivalent. If you want to continue with analysis, combine these 3 components into a single matrix.

```{r combine matrices}
comb.dist <- (dist[,,3] + 
                dist[,,2] +
                dist[,,1])
comb.dist[which(is.na(comb.dist))] <- 0 
comb.dist <- comb.dist + t(comb.dist) # reflect the data so it's easier to look up distances
```

### 2.3 Visualize distances
We can visualize the distances between the species using Principal Coordinates Analysis:

```{r PCoA plot}
PCoA <- cmdscale(comb.dist, eig=TRUE, k=2)

library('ggplot2')
pts <- data.frame(PCoA$points)
dimnames(pts)[[2]] <- c("PCoA 1","PCoA 2")
pts$species <- dimnames(pts)[[1]]

ggplot(data = pts, aes(x = `PCoA 1`, y = `PCoA 2`)) +
  geom_point() +
  geom_label(aes(label = species), position = position_nudge(x = 0, y = 0.5))

```
